--- 
title: "m6a analysis"
author: "tsztank"
date: "`r format(Sys.time(), '%d/%B/%Y')`"
output: 
  html_document:
    theme: journal
---

```{r knitr_options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  cache = T,
  autodep = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = NA,
  cache.path = 'cache/',
  fig.path = 'figures/')
```

```{r load_libraries, echo = F, cache = F}
library(plyr)
library(tidyverse)
library(biomaRt)
library(magrittr)
library(yaml)
library(readr)
library(cowplot)
library(wesanderson)
library(GGally)
library(DESeq2)
library(tximport)
```

## Sanity checks

The if we plot the counts and tpm-s (transcript per million) on a log-x scale, we observe a normal distribution which is expected. Also, the mean values of tpm-s are comparable, so it seems there is not big difference between the libraries of the 3 conditions sequenced.

```{r get_gene_name_id_map, echo = F}
ensembl = useMart('ensembl', dataset = 'dmelanogaster_gene_ensembl')

genes <- ensembl %>%
	getBM(
		  attributes = c('ensembl_gene_id',
						 'ensembl_transcript_id',
						 'external_gene_name',
						 'external_transcript_name'),
		  mart = .) %>%
	as_tibble

colnames(genes) <- c('gene_id', 'tx_id', 'gene_name', 'tx_name') 
```

```{r load_kallisto_results, echo = F}
config = read_yaml('../../config.yaml')
samples <- config$samples

kallisto_dir = '../../data/kallisto_quant/'

kallisto_data <- samples %>% ldply(function(smpl){
		df <- read_table2(paste0(kallisto_dir, '/', smpl, '/abundance.tsv'))
		df$sample <- smpl

		df
	}) %>% 
	as_tibble %>%
	separate(sample, into = c('condition', 'replicate'), sep = '[_]') %>%
	dplyr::rename(tx_id = target_id) %>%
	mutate(tx_id = as.factor(tx_id),
		   condition = as.factor(condition)) %>%
	left_join(genes, by = 'tx_id')
```

```{r plot_corr_between_runs, echo = F, eval = F}
kallisto_data %>%
	unite('run', condition, replicate) %>%
	dplyr::select(tx_id, run, tpm) %>%
	spread(run, tpm) %>%
	{
		ggpairs(., columns = 2:ncol(.))
	}
```

```{r plot_histogram_tpm, echo =F}
kallisto_data %>%
	ggplot(aes(tpm, fill = condition)) +
	geom_density(alpha = 0.8) +
	scale_x_log10() +
	scale_fill_manual(values=wes_palette(n=3, name="GrandBudapest1")) +
	ggtitle('TPM comparison between conditions')
```

```{r plot_histogram_counts, echo =F}
kallisto_data %>%
	ggplot(aes(est_counts, fill = condition)) +
	geom_density(alpha = 0.8) +
	scale_x_log10() +
	scale_fill_manual(values=wes_palette(n=3, name="GrandBudapest1")) +
	facet_grid(condition~replicate) +
	theme(axis.text.x = element_text(angle = 45, hjust =1)) +
	ggtitle('Estimated counts of each run')
```

## Looking at genes of interest

First lets look at 3 genes: Clk, tim and vri.

It seems that for these 3 genes, IgG\_3 is an outlier.

```{r plot_genes_of_interest, echo = F}
genes_of_interest <- c('vri', 'Clk', 'tim')

plot_genes <- function(kallisto_data, genes){
	kallisto_data %>%
		filter(gene_name %in% genes) %>%
		group_by(gene_name, condition, replicate) %>%
		dplyr::summarise(tpm = sum(tpm)) %>%
		ggplot(aes(gene_name, tpm, colour = condition, shape = replicate)) +
			geom_point(size = 4, position = position_dodge(width=0.3)) +
			scale_colour_manual(values=wes_palette(n=3, name="GrandBudapest1"))
}

plot_genes(kallisto_data, genes_of_interest)
```

Lets look at Mbl separately. Here we again can observe that IgG\_3 is an outlier compared to others.

```{r plot_mbl_tpm, echo = F}
plot_genes(kallisto_data, c('mbl'))
```

```{r plot_correlation_between_replicates, echo = F, eval =F}
get_cor_plot <- function(kallisto_data, sx, sy){
	dat <- kallisto_data %>%
		unite('run', condition, replicate) %>%
		dplyr::select(tx_id, run, tpm) %>%
		spread(run, tpm)

	r <- cor(dat[[sx]], dat[[sy]])

	pl <- dat %>%
		# convert tpm to log for visualization
		gather('run', 'tpm', -tx_id) %>%
		mutate(tpm = log10(tpm)) %>%
		spread(run, tpm) %>%
		ggplot(aes_string(x=sx, y=sy)) +
			geom_point()

	pl <- pl + annotate("text", Inf, -Inf, label = paste0('R = ', round(r, 3)), hjust = 2, vjust = -10)

	pl
}

get_cor_plot(kallisto_data, 'IgG_3', 'IgG_2')
```


```{r prepare_deseq_de_analysis, echo = F}
tx2gene <- genes %>%
	transmute(TXNAME = tx_id,
			  GENEID = gene_id)

kallisto_metadata <- tibble(sample = samples) %>%
	separate(sample, into = c('condition', 'replicate', 'rnd'), remove = F) %>%
	# remove randomly generated suffix
	dplyr::select(-rnd) %>%
	mutate(path = paste0(kallisto_dir, sample),
		   condition = factor(condition, levels = c('input', 'IgG', 'IP'))) %>%
	# sort tibble by factor
	arrange(condition)

txi <- kallisto_metadata %>%
	mutate(files = paste0(path, '/abundance.tsv')) %>%
	{
		files <- .$files
		names(files) <- .$sample 
		tximport(files, type = 'kallisto', tx2gene = tx2gene, ignoreAfterBar = T)
	}

dds <- DESeqDataSetFromTximport(txi, as.data.frame(kallisto_metadata), ~condition)
```

### PCA analysis

From the below PCA plot we can clearly see, that IgG\_3 is an outlier. For further analysis, it is therefore probably better to not take it into account.

```{r pca_analysis, echo = F}
vsd <- vst(dds, blind=FALSE)
pca <- plotPCA(vsd, intgroup=c('condition', 'replicate'), returnData=T) %>%
	as_tibble

pca %>%
	# modify factor level to have consistent colors
	mutate(condition = factor(condition, levels = c('IgG', 'input', 'IP'))) %>%
	ggplot(aes(PC1, PC2, colour = condition, shape = replicate)) + 
	geom_point(size = 4) +
	scale_colour_manual(values=wes_palette(n=3, name="GrandBudapest1"))

```

## TODO

- Run DESeq2 Differential Expression pipeline to check for enriched genes in IP.

